---
title: "Exploring the Palmer Penguins Dataset"
author: "Parva Fazeli"
date: "2026-09-24"
---

## Introduction

The Palmer Penguins dataset contains measurements collected from three penguin species in the Palmer Archipelago, Antarctica: Adelie, Chinstrap, and Gentoo. It includes several physical measurements for individual penguins, such as body mass, flipper length, bill length, and bill depth, as well as information about species, sex, island, and year.

In this post, I use R to explore two questions:

1. How does body mass differ among the three penguin species?
2. What is the relationship between flipper length and body mass?

The analysis includes basic data exploration, handling of missing values, numerical summaries, and visualizations.

## Loading the packages

First, I load the packages required for the analysis.

```{r}
#| message: false
#| warning: false

library(palmerpenguins)
library(tidyverse)
```

## Data

The data used in this analysis come from the [Palmer Penguins](https://allisonhorst.github.io/palmerpenguins/) dataset from Palmer Station Antarctica LTER.

The `palmerpenguins` package provides the dataset, while functions from the `tidyverse` are used for data manipulation and visualization.

## Exploring the data

Before beginning the analysis, I inspect the first few rows of the dataset.

```{r}
head(penguins)
```

I also examine the dimensions of the dataset.

```{r}
dim(penguins)
```

Each row represents an individual penguin, while the columns represent different variables measured or recorded for that penguin.

The variables that are particularly important for this analysis are:

- `species`: penguin species
- `body_mass_g`: body mass measured in grams
- `flipper_length_mm`: flipper length measured in millimetres

## Handling missing values

Before creating the summaries and visualizations, I check for missing values in the two numerical variables used in this analysis.

```{r}
penguins |>
  summarise(
    missing_body_mass = sum(is.na(body_mass_g)),
    missing_flipper_length = sum(is.na(flipper_length_mm))
  )
```

Because the analysis depends on both body mass and flipper length, I create a cleaned dataset containing only observations for which both measurements are available.

```{r}
penguins_clean <- penguins |>
  drop_na(body_mass_g, flipper_length_mm)
```

I can compare the number of observations before and after cleaning.

```{r}
nrow(penguins)
nrow(penguins_clean)
```

Removing these incomplete observations allows the same cleaned dataset to be used consistently throughout the analysis.

## Question 1: How does body mass differ among penguin species?

The first question examines how body mass differs among the three penguin species.

I begin by calculating several summary statistics for each species.

```{r}
body_mass_summary <- penguins_clean |>
  group_by(species) |>
  summarise(
    n = n(),
    mean_body_mass = mean(body_mass_g),
    median_body_mass = median(body_mass_g),
    sd_body_mass = sd(body_mass_g),
    min_body_mass = min(body_mass_g),
    max_body_mass = max(body_mass_g)
  )

body_mass_summary
```

The mean and median describe the typical body mass for each species. The standard deviation describes how much body mass varies within each species, while the minimum and maximum show the observed range.

### Visualizing body mass by species

A boxplot provides another way to compare the body-mass distributions of the three species.

```{r}
ggplot(
  penguins_clean,
  aes(
    x = species,
    y = body_mass_g,
    fill = species
  )
) +
  geom_boxplot() +
  labs(
    title = "Body Mass of Palmer Penguins by Species",
    x = "Species",
    y = "Body Mass (g)"
  ) +
  theme_minimal() +
  guides(fill = "none")
```

### Interpretation

The numerical summaries and boxplot show clear differences in body mass among the three species. Gentoo penguins tend to have greater body masses than Adelie and Chinstrap penguins. Adelie and Chinstrap penguins have more similar body-mass distributions, although body mass still varies within each species.

The summary statistics provide numerical information about the centre, variability, and range of body mass, while the boxplot makes it easier to visually compare the distributions across species.

## Question 2: What is the relationship between flipper length and body mass?

The second question investigates whether penguins with longer flippers also tend to have greater body mass.

First, I calculate the overall correlation between the two numerical variables.

```{r}
penguins_clean |>
  summarise(
    correlation = cor(flipper_length_mm, body_mass_g)
  )
```

The correlation provides a numerical measure of the direction and strength of the relationship between flipper length and body mass.

I also calculate the correlation separately for each species.

```{r}
penguins_clean |>
  group_by(species) |>
  summarise(
    correlation = cor(flipper_length_mm, body_mass_g)
  )
```

Calculating the correlation separately for each species helps determine whether the relationship between flipper length and body mass is also present within individual species, rather than being driven only by differences among species.

### Visualizing the relationship

Because both variables are numerical, a scatterplot is useful for examining their relationship.

```{r}
ggplot(
  penguins_clean,
  aes(
    x = flipper_length_mm,
    y = body_mass_g,
    color = species
  )
) +
  geom_point(alpha = 0.7) +
  labs(
    title = "Relationship Between Flipper Length and Body Mass",
    subtitle = "Palmer Penguins",
    x = "Flipper Length (mm)",
    y = "Body Mass (g)",
    color = "Species"
  ) +
  theme_minimal()
```

### Interpretation

The scatterplot shows a positive relationship between flipper length and body mass. In general, penguins with longer flippers also tend to have greater body mass.

The species groups are also visible in the plot. Gentoo penguins generally appear toward the upper-right portion of the graph, corresponding to longer flippers and greater body mass. Adelie and Chinstrap penguins tend to have shorter flippers and lower body masses.

The species-specific correlations provide additional information about the relationship within each species. Comparing these values with the overall correlation helps distinguish the within-species relationships from the overall pattern observed when all three species are considered together.

## Conclusion

This analysis used the Palmer Penguins dataset to investigate differences in body mass among species and the relationship between flipper length and body mass.

Before conducting the analysis, observations with missing body-mass or flipper-length measurements were removed so that the same cleaned dataset could be used consistently throughout the post.

For the first question, the numerical summaries and boxplot showed that body mass differs across species, with Gentoo penguins generally having greater body mass than Adelie and Chinstrap penguins. The summary statistics also provided information about the centre, variability, and range of body mass within each species.

For the second question, the overall correlation and scatterplot indicated a positive relationship between flipper length and body mass. Penguins with longer flippers generally tended to be heavier. Calculating the correlation separately for each species also allowed this relationship to be examined within each species rather than only across the complete dataset.